Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device

### 2023-07-07 manuscript/manuscript.tex
results almost done. XXgboost plots are ok.

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{soul}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{npj Digital Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device},
  pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
  pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
quality metrics},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality metrics
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality metrics, including sleep period time
(SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO). \st{By
analyzing these additional metrics, we aimed to provide a comprehensive
evaluation of the model's capability in assessing various aspects of
sleep quality, allowing us to determine the overall effectiveness of our
models in accurately estimating different parameters related to sleep
duration and quality.}

@@

The current study leverages data from the SCREENS
project\citep{rasmussen2020}, a study conducted from October 2018 to
March 2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we isolated
data from child participants between the ages of 6 to 10 years within
the SCREENS cohort. Our main sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, utilizes advanced EEG hardware and signal
processing algorithms, employs three self-adhesive, disposable sensors
placed outside the hairline for reliable EEG signal acquisition. The ZM
uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is utilized
for accurate sleep detection, showcasing its aptness for in-home
monitoring\citep{kaplan2014}, while the Z-PLUS effectively
differentiates sleep stages, as evidenced by its alignment with expert
evaluations using PSG data\citep{wang2015}.

Finally, we affirm that the SCREENS study received approval from the
Regional Scientific Committee of Southern Denmark, and all data handling
processes complied with the General Data Protection Regulation (GDPR),
guaranteeing the ethical and secure management of participant
information.

@@

In this study, raw data processing began with a low-pass filtration step
using a 4th order Butterworth filter with a 5 Hz cut-off frequency to
eliminate high-frequency noise. Following filtration, data were
partitioned into overlapping 2-second intervals, each successive
interval sharing a 50\% overlap with the previous one similar to methods
described by Skotte et al.\citep{skotte_detection_2014}. Any non-wear
data was remove using previously described methods\citep{skovgaard2023}
and data was resampled to 30-second epochs so every sample classified by
the algorithms corresponds to a 30-second epoch scored during the ZM
recordings. Subsequently, we performed a feature extraction process that
yielded a set of 88 features, providing a robust characterization of the
data. Extracted from accelerometer and temperature signals, these
features include temporal elements that use both lag and lead values,
capturing dynamic data trends by incorporating measurements from
preceding and upcoming intervals. Furthermore, inspired by Walch et
al.\citep{walch2019}, we developed sensor-independent features to
encapsulate circadian rhythms. These features offer unique insights not
directly discernible from sensor outputs and are meant to approximate
the changing drive of the circadian clock to sleep over the course of
the night (see Figure~\ref{fig-sensor-independent}). Furthermore, the
feature set was enriched by including signal characteristics, which
encompass vector magnitude, mean crossing rate, skewness, and kurtosis
for each of the x, y, and z dimensions. These features provide a
comprehensive understanding of the signal's attributes, facilitating a
more nuanced analysis of the underlying patterns. All features are
summarized in table ??? (overvej tabel til supp. material)

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings commonly occurring at the end of each
cycle\citep{galland_normal_2012}. Examining the raw ZM predictions, we
noted a significant overestimation in the number of awakenings per night
for the children in our study, exceeding what would be expected based on
typical sleep cycle patterns. Consequently, we elected to train and
evaluate our models using not only the raw ZM output, but also versions
that were subjected to 5-minute and 10-minute median filters. This
approach resulted in an anticipated, more age-appropriate count of
awakenings per night, providing a more accurate depiction of children's
sleep patterns (see Figure~\ref{fig-zm-median}).

@@

In our study, we utilize two distinct modeling strategies to predict
sleep patterns derived from thigh-mounted accelerometer data: a
sequential ensemble and a multiclass bidirectional Long Short-Term
Memory (biLSTM)\citep{hochreiter1997} neural network. The sequential
ensemble involves using four pairs of models, each pair consisting of
the same algorithm, that predict in-bed and sleep time successively.
Conversely, the biLSTM is a comprehensive model that predicts multiple
classes simultaneously, effectively handling complex temporal patterns.
By applying and comparing both, we aim to evaluate their relative
effectiveness.

@@

The biLSTM is a multiclass classifier capable of predicting the classes
of out-bed-awake, in-bed-awake, and in-bed-asleep. The architecture of
the biLSTM consisted of four layers, each with 128 hidden units. The
choice of four layers and 128 hidden units in our model balanced
complexity and efficiency: it was deep enough to learn intricate
patterns yet feasible to train timely. Additionally, the bidirectional
nature of the LSTM, doubling the hidden units at each time step,
improved data comprehension and avoided overfitting.

something on sequence length and step size

The biLSTM model was chosen for sleep classification using accelerometer
data from a thigh-mounted device due to its advantages in modeling
temporal context, handling a large temporal scope, and enabling temporal
inference over any feature without the need for hand-designed temporal
features.

Previous studies have explored the use of LSTM models for sleep
detection, showing promising results in capturing complex temporal
patterns. Notable works by Sano et al. \citep{sano2019} and Chen et al.
\citep{chen2021} have demonstrated the potential of LSTM models in
improving sleep detection based on accelerometer data. However, further
research is necessary to fully comprehend the strengths and limitations
of this approach.

@@

For the models in sequence, we trained four pairs of classification
models, each pair distinguishing between in-bed/out-of-bed and
asleep/awake states. The dataset was randomly split into a training set
(approximately 50\% of the subjects) and a testing set (also
approximately 50\%), ensuring that samples from the same subject were
never simultaneously used in both sets. Hyper-parameter optimization was
performed using 10-fold Monte Carlo cross-validation, with the F1 score
as the performance metric. The best-performing set of hyperparameters
was then used to fit the models to the full training dataset, maximizing
accuracy by leveraging all available data. Furthermore, the dataset for
predicting sleep was highly imbalanced which can pose challenges when
training machine learning models, as models may favor predicting the
majority class. To address this issue, we employed oversampling
techniques to ensure that each class had a roughly equal number of
training samples. Specifically, we utilized the Synthetic Minority
Over-sampling Technique (SMOTE)\citep{chawla2002}, which generates new
samples by interpolating random samples with their nearest neighbors. In
our study, we implemented SMOTE using the themis R
package\citep{themis}, resampling all classes to achieve a balanced
distribution of training samples.

The biLSTM model was trained using the Adam optimizer, which is
computationally efficient and adapts the learning rate during training.
Cross-entropy loss function was employed for its suitability in
multiclass classification with mutually exclusive classes. The softmax
activation function was selected for the output layer to obtain a
probability distribution over the classes. Data for training the biLSTM
were randomly divided into a training, validation, and test tensors
based on a 50/25/25 split. The model was evaluated using the F1 score on
both the training and validation sets. Early stopping was implemented
with a patience of 3 epochs, meaning training would stop if there was no
improvement in the validation loss for 3 consecutive epochs.

\hypertarget{model-validation-evt.-bare-statistics}{%
\subsubsection{Model Validation {[}evt. bare
``statistics''{]}}\label{model-validation-evt.-bare-statistics}}

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs.~out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs.~awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. Furthermore, we
present precision-recall curves and receiver operating characteristic
curves, including the area under the curve, for each model and for each
type of ZM prediction.We considered both the in-bed/out-of-bed and
awake/asleep scoring tasks as binary classification problems,
designating in-bed and asleep as the positive labels and out-of-bed and
awake as the negative labels in accordance with previous
research\citep{hjorth2012, kushida2001}.

To evaluate our developed models' performance in assessing sleep quality
metrics, we employed Bland-Altman plots and Pearson correlations.
Additionally, we used the Bland-Altman method to assess the agreement
between two measurement techniques. Given the multiple observations per
subject in our dataset, we incorporated a bootstrap procedure to account
for this additional source of variability. We computed the mean
difference (bias) and the limits of agreement as the mean difference
±1.96 times the standard deviation of the differences. The 95\%
confidence intervals for the bias and the limits of agreement were
estimated using the bias-corrected and accelerated bootstrap method with
10,000 bootstrap replicates, to ensure our measurements are robust and
account for intra-subject variability. These methods were used to
compare the sleep quality summaries calculated by our models with those
obtained from EEG-based ZM sleep quality summaries. The sleep quality
summaries included the following metrics:

@@

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{npj Digital Medicine}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device},
  pdfauthor={Esben Høegholm Lykke; Jan Christian Brønd},
  pdfkeywords={Sleep, Accelerometry, EEG, Machine learning, Sleep
quality},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality parameters, including sleep period
time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO).

@@

The current study leverages data from the SCREENS
project\citep{rasmussen2020}, a study conducted from October 2018 to
March 2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we focused on
data from child participants aged between 6 and 10 years within the
SCREENS cohort. Our primary sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
utilized for accurate sleep detection, showcasing its suitability for
in-home monitoring\citep{kaplan2014}, while the Z-PLUS effectively
differentiates sleep stages, as evidenced by its alignment with expert
evaluations using PSG data\citep{wang2015}. In the current study, we
treated all sleep stages as a single category effectively deducing the
output of the ZM to ``awake'' and ``asleep'' as the ability to
distinguish sleep stages are not a necessity to derive sleep quality
parameters of interest and to simplify the learning process of the
models.

Figure~\ref{fig-flow} illustrates the selection criteria applied to the
children's recordings from the SCREENS study. We included only ZM
recordings that were accompanied by complete accelerometer data and
lasted between 7 and 14 hours. Any night when the ZM reported sensor
issues was excluded. The children whose recordings were considered had
an average age of 9.4 years, with a standard deviation of 2.1. In their
raw form, the ZM predictions encompassed 696,779 epochs, each 30 seconds
long. Notably, approximately 84\% of the total ZM recording duration was
classified as sleep, resulting in an imbalance of the dataset.

\caption{\label{fig-flow}Flowchart of eligible ZM recording nights
included in the study}

Finally, we affirm that the SCREENS study received approval from the
Regional Scientific Committee of Southern Denmark, and all data handling
processes complied with the General Data Protection Regulation (GDPR),
ensuring the ethical and secure management of participant information.

@@

In this study, data processing of the raw accelerometer data began with
a low-pass filtration step using a 4th order Butterworth filter with a 5
Hz cut-off frequency to eliminate high-frequency noise. Following
filtration, data were partitioned into overlapping 2-second intervals,
each successive interval sharing a 50\% overlap with the previous one
similar to methods described by Skotte et
al.\citep{skotte_detection_2014}. Any non-wear data was remove using
previously described methods\citep{skovgaard2023} and data was resampled
to 30-second epochs so every sample classified by the algorithms
corresponds to a 30-second epoch scored during the ZM recordings.
Subsequently, we performed a feature extraction process that yielded a
set of 88 features, providing a robust characterization of the data.
Extracted from accelerometer and temperature signals, these features
include temporal elements that use both lag and lead values, capturing
dynamic data trends by incorporating measurements from preceding and
upcoming intervals. Furthermore, inspired by Walch et
al.\citep{walch2019}, we incorporated sensor-independent features to
encapsulate circadian rhythms. These features offer unique insights not
directly discernible from sensor outputs and are meant to approximate
the changing drive of the circadian clock to sleep over the course of
the night (see Figure~\ref{fig-sensor-independent}). Furthermore, the
feature set was enriched by including signal characteristics, which
encompass vector magnitude, mean crossing rate, skewness, and kurtosis
for each of the x, y, and z dimensions. All features are summarized in
table ??? \textbf{in the supplementary matrials}. Subsequently, we
merged the ZM and corresponding accelerometer recordings. Any
overlapping time between the ZM and accelerometer data was treated as
`in-bed' time, with the remaining time considered `out-of-bed'. This
process yielded a dataset providing a around the clock temporal view of
each participant's activity and sleep patterns.

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings most likely occurring at the end of each
cycle\citep{galland_normal_2012}. Upon examining the raw ZM predictions,
we noted a significant overestimation in the number of awakenings per
night for the children in our study, exceeding what would be expected
based on typical sleep cycle patterns. In particular, many of these
brief awakenings could be considered as noise, which when present in the
data, can potentially hinder the learning process of machine learning
models by obscuring the underlying patterns that the models are trying
to learn, leading to less accurate predictions. Consequently, we elected
to train and evaluate our models using not only the raw ZM output, but
also versions that were subjected to 5-minute and 10-minute median
filters. This approach, by mitigating this noise, resulted in an
anticipated, more age-appropriate count of awakenings per night,
providing a more accurate depiction of children's sleep patterns (see
Figure~\ref{fig-zm-median}).

@@

In this study, we employed two distinct modeling strategies to analyze
sleep patterns from thigh-mounted accelerometer data. We used a
sequential strategy, comprising an ensemble of four pairs of models,
each pair featuring the same algorithm. This strategy aimed to make the
prediction task more manageable for the algorithms by breaking it down
into a sequence of two binary classifications: first predicting `in-bed'
time, then `sleep' time. Simultaneously, we also used a multiclass
approach utilizing a bidirectional Long Short-Term Memory
(biLSTM)\citep{hochreiter1997} neural network.

@@

In this study, we employed a biLSTM, a multiclass classifier, to predict
three distinct states: out-of-bed-awake, in-bed-awake, and
in-bed-asleep. The architecture of the biLSTM was set up with four
layers, each equipped with 128 hidden units. This configuration was
intentionally chosen to balance between model complexity and training
efficiency: it provided the depth necessary for learning intricate
patterns while remaining feasible for timely training. The bidirectional
design of the LSTM served to enhance data interpretation and mitigate
overfitting by doubling the hidden units at each time step. For input,
we used tensors shaped as sequences, with each sequence spanning 10
minutes and a step size of one.This approach follows in the footsteps of
previous studies that utilized LSTM models for sleep detection. These
studies showcased the promising potential of LSTMs in capturing complex
temporal patterns. Particularly, the works of Sano et al.
\citep{sano2019} and Chen et al. \citep{chen2021} demonstrated the
effectiveness of LSTM models in improving sleep detection using
accelerometer data, underscoring the value of this modeling approach.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately
50\% of the subjects. This division ensured that samples from the same
night were never simultaneously present in both sets. To optimize
hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
a regular grid, comprising 20 different combinations of hyperparameters.
The F1 score served as the optimization metric. The best-performing set
of hyperparameters was then used to fit the models to the full training
dataset. This approach allowed us to maximize performance by leveraging
all available data. Moreover, after extracting the in-bed time from the
initial sequential models, the imbalance on the resulting dataset could
cause biases during model training, as models may favor predicting the
majority class. To rectify this, we employed the Synthetic Minority
Over-sampling Technique (SMOTE)\citep{chawla2002}. SMOTE generates new
samples by interpolating random samples with their nearest neighbors. We
utilized the themis R package\citep{themis} to implement SMOTE,
resulting in a balanced distribution of training samples across both
classes.

The biLSTM model was trained to differentiate between three states:
out-of-bed-awake, in-bed-awake, and in-bed-asleep. The data used for
training the biLSTM was randomly divided into training, validation, and
test sets, based on a 50/25/25 split. We ensured that data from the same
night was not present across different sets. The model was trained using
the Adam optimizer, selected for its computational efficiency and
adaptability of the learning rate during training. Given the multiclass
classification task with mutually exclusive classes, we employed the
cross-entropy loss function. To obtain a probability distribution over
the classes, the softmax activation function was applied at the output
layer. We evaluated the model's performance using the F1 score on both
the training and validation sets. We implemented early stopping with a
patience of 3 epochs, halting the training process if there was no
improvement in the validation loss over three consecutive epochs.

\hypertarget{model-validation}{%
\subsubsection{Model Validation}\label{model-validation}}

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs.~out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs.~awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. To further
illustrate model performance, we provide confusion matrices for the full
dataset, encompassing both in-bed and out-of-bed data. These matrices
report relative counts, column percentages (the proportion of the true
class accurately predicted), and row percentages (the proportion of
predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research\citep{hjorth2012, kushida2001}.

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the limits of agreement as the mean
difference plus or minus 1.96 times the standard deviation of these
differences. To ensure our measurements were robust and accounted for
intra-subject variability, we estimated the 95\% confidence intervals
for both the bias and the limits of agreement using a bias-corrected and
accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
sleep quality parameteres included the following:

@@

As reported in Table~\ref{tbl-zm_overview} the sleep quality parameters
derived from ZM predictions were modified by the implementation of
5-minute and 10-minute median filters. SPT were consistent across raw
and filtered datasets (mean: 9.2 ± 2.1 hours), corresponding to the
length of the ZM recording. TST and SE increased in the filtered data,
implying the filters categorize some wakefulness as sleep. Specifically,
TST increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively. LPS
also elevated, suggesting the filter smooths out brief awakenings at
sleep onset, leading to a prolonged time to persistent sleep. A
significant change was seen in WASO, which dropped from 39.0 ± 33.6
minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
the 5-minute and 10-minute filtered data, respectively. The number of
awakenings was also considerably reduced with the application of
filters. In the raw data, the average number of awakenings was 34.46 ±
11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
5-minute and 10-minute filtered data sets respectively.

\toprule
 & SPT (hrs) & TST (hrs) & SE (\%) & LPS (min) & WASO (min) & Awakenings (N) \\ 
\midrule
Raw ZM Predictions & 9.2 (2.1) & 7.7 (1.9) & 82.6 (12) & 34.5 (27.9) & 39 (33.6) & 34.5 (11.3) \\ 
5-Min Median & 9.2 (2.1) & 8.1 (2) & 86.4 (12.7) & 36.3 (39.8) & 30.6 (46.8) & 4.4 (3.3) \\ 
10-Min Median & 9.2 (2.1) & 8.2 (2.1) & 87.5 (12.9) & 38 (48.7) & 22.3 (55.4) & 1.9 (2) \\ 
\bottomrule
\end{longtable}

\hypertarget{performance-on-epoch-to-epoch-basis}{%
\subsection{Performance on Epoch-to-Epoch
Basis}\label{performance-on-epoch-to-epoch-basis}}

The epoch-to-epoch evaluation of predicting in-bed time is outlined in
Table~\ref{tbl-in_bed_performance}, and demonstrates practically
equivalent performance across all model types. The F1 score ranges from
94.4\% (Decision Tree) to 95.4\% (XGBoost), while accuracy ranges from
95.3\% (Decision Tree) to 96.1\% (XGBoost). Sensitivity, Precision, and
Specificity also demonstrate consistent results across the models. The
XGBoost model, despite recording the highest metrics with an F1 score of
95.4\% and accuracy of 96.1\%, outpaced the others only marginally.

\toprule
 & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
\midrule
Decision Tree & $94.4$ & $95.3$ & $93.1$ & $95.6$ & $96.9$ \\ 
Logistic Regression & $95.0$ & $95.7$ & $95.0$ & $94.9$ & $96.3$ \\ 
Feed-Forward Neural Net & $95.0$ & $95.8$ & $95.1$ & $95.0$ & $96.3$ \\ 
XGBoost & $95.4$ & $96.1$ & $95.8$ & $94.9$ & $96.2$ \\ 
\bottomrule
\end{longtable}

Table~\ref{tbl-sleep_performance} details the performance of the second
sequential models on raw and median-filtered (5 and 10 minute) ZM
predictions. These are the predictions of the sequential models' ability
to predict asleep/awake on only the extracted in-bed time. The F1
scores, which are unweighted macro averages, for raw ZM predictions
range from 71.05\% to 76.18\%. The models perform comparably, but the
low specificity values (62.84\% to 70.93\%) suggest difficulty in
correctly classifying awake epochs. Applying 5-minute median filtering
improves the performance metrics. The XGBoost model tops the charts with
an F1 score of 79.22\% and NPV of 74.00\%. However, specificity still
remains low, with values between 54.68\% and 74.84\% across all models.
With 10-minute median filtering, the metrics improve further. The
XGBoost model still leads with an F1 score of 80.87\% and an NPV of
75.76\%. But, specificity remains low, ranging from 57.47\% to 76.35\%
across all models.

\hypertarget{tbl-sleep_performance}{}
\begin{longtable}{lrrrrr}
\caption{\label{tbl-sleep_performance}Sleep Performance Metrics }\tabularnewline
